Reformulating phylogenetic mixed models to improve flexibility and speed

### 2020-01-20 phyloglmm_ms.tex
forgot to merge with BMB's recent changes, merging now

Standard statistical regression techniques neglect correlations among species responses due to their shared evolutionary history.
Classic phylogenetic regression uses a statistical model in which phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the dependent-variable trait evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}. 
If in addition the residuals are normally distributed and observed without any additional error (e.g., from other covariates \bmb{???}) or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies, nicolakakis2000forebrain} is sufficient to account for the phylogenetic correlation. \bmb{what's the second citation for?}
More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary processes other than the simple Brownian motion model. 
These methods partition residual variation into two components: (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
If each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic variation and tip variation can both be considered part of the evolutionary process error (which we will call tip variation or intercept-level variation) while the among species residual variation is associated with among observation variation within each species.
%\bmb{This is a good point, well expressed. Does PGLMM lump tip variation and within-species variation, or is this primarily an issue with Blomberg/Pagel? Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots
%\mli{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely.
Indeed, many existing methods are restricted to single observations per species, requiring users to collapse their observations to mean values for each species, which can be limiting if some covariates of interest vary within species.
%% REMEMBER TO REVISIT IN DISCUSSION

@@

%% FIXME: chart looks ugly
\begin{table}[]
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{Model} & \textbf{Method} & \textbf{Data} & \textbf{Platform} \\ \hline
\multirow{2}{*}{
\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Model (GLM)
\end{tabular}} & Correlated residual & Single observation & \begin{tabular}[c]{@{}l@{}}\pkg{nlme}:gls, \\ \pkg{ape}:pic\end{tabular} \\ \cline{2-4} 
& \begin{tabular}[c]{@{}l@{}}Residual \\ + phylogenetic intercept\end{tabular} & Single observation & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single observation\\ Balanced design\end{tabular} & \pkg{pez}                                                                                            \\ \cline{3-4} 
                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{lme4}, \pkg{glmmTMB}, \pkg{phyr}                                                                                  \\ \hline
\multirow{2}{*}{Bayesian GLMM}                                                                    & \multirow{2}{*}{Random effect}                                        & Balanced design                                                              & \pkg{MCMCglmm}                                                                                       \\ \cline{3-4} 
                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{brms}                                                                                           \\ \hline
\end{tabular}
\caption{GLM-based phylogenetic comparative methods and their implementations in R packages.
\bmb{can we add ``per species'' to ``single observation'' without wrecking the table?}}
\label{table:model}
\end{table}

>>>>

Standard statistical regression techniques neglect correlations among species responses due to their shared evolutionary history.
Classic phylogenetic regression uses a statistical model in which phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the dependent-variable trait evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}. 
If in addition the residuals are normally distributed and observed without any additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies} is sufficient to account for the phylogenetic correlation.
More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary processes other than the simple Brownian motion model. 
These methods partition residual variation into two components: (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
If each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic variation and tip variation can both be considered part of the evolutionary process error (which we will call tip variation or intercept-level variation) while the among species residual variation is associated with among observation variation within each species.
%\bmb{This is a good point, well expressed. Does PGLMM lump tip variation and within-species variation, or is this primarily an issue with Blomberg/Pagel? Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots
%\mli{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely.
Indeed, many existing methods are restricted to single observations per species, requiring users to collapse their observations to mean values for each species, which can be limiting if some covariates of interest vary within species.
%% REMEMBER TO REVISIT IN DISCUSSION

@@

%% FIXME: chart looks ugly
\begin{table}[]
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{Model} & \textbf{Method} & \textbf{Data} & \textbf{Platform} \\ \hline
\multirow{2}{*}{
\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Model (GLM)
\end{tabular}} & Correlated residual & \begin{tabular}[c]{@{}l@{}}Single observation \\ per species \end{tabular} & \begin{tabular}[c]{@{}l@{}}\pkg{nlme}:gls, \\ \pkg{ape}:pic\end{tabular} \\ \cline{2-4} 
& \begin{tabular}[c]{@{}l@{}}Residual \\ + phylogenetic intercept\end{tabular} & \begin{tabular}[c]{@{}l@{}}Single observation \\ per species \end{tabular} & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single observation \\per species, \\ Balanced design\end{tabular} & \pkg{pez}                                                                                            \\ \cline{3-4} 
                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{lme4}, \pkg{glmmTMB}, \pkg{phyr}                                                                                  \\ \hline
\multirow{2}{*}{Bayesian GLMM}                                                                    & \multirow{2}{*}{Random effect}                                        & Balanced design                                                              & \pkg{MCMCglmm}                                                                                       \\ \cline{3-4} 
                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{brms}                                                                                           \\ \hline
\end{tabular}
\caption{GLM-based phylogenetic comparative methods and their implementations in R packages.
\bmb{can we add ``per species'' to ``single observation'' without wrecking the table?}}
\label{table:model}
\end{table}
